---
title: "apoE_cmd_github_240814_CL.Rmd"
author: "Caroline Lindblad"
date: "2024-07-14"
output: html_document
---

META INFORMATION 
This code underlies the paper **Influence of Apolipoprotein E genotype on the proteomic profile in cerebral microdialysis after human severe TBI: a prospective observational study**. 

All code was written by the first author of the paper (Caroline Lindblad).

Further questions should be directed to:
caroline.lindblad@ki.se
caroline.i.lindblad@gmail.com
caroline.lindblad@uu.se
or via telephone
+46708-236282


# Wed Aug 14 18:41:14 2024 ------------------------------


```{r}

rm(list = ls())

```


```{r}

library(gridExtra)

```


import sourced script
```{r}

source("./scripts/script_apoe/idHarmonization_APOE_230828_CL.R")

#the sourced script imports numerous other libraries including tidyverse, ggplot2 
#the sourced script contains data cleaning operations and preparatory data operations for inferential analysis

```

filter the data so that only one allele is presented (we are only interested in genotype)
```{r}

vars_to_remove <- c("apoEhaplotype", "allele_number")

apoEstatus_idDf_protq_olink <- apoEstatus_idDf_protq_olink[, !names(apoEstatus_idDf_protq_olink) %in% vars_to_remove]

apoEstatus_idDf_protq_olink <- distinct(apoEstatus_idDf_protq_olink) #the data is reduced to not demonstrate both alleles (thus only genotype)

```

cols of interest for dimensionality reduction techniques
c(15:106)

```{r}

olink_only <- apoEstatus_idDf_protq_olink[, c(1:8, 10:11, 15:106, 109)] #new 231008

olink_only <- olink_only %>%
  pivot_wider(names_from = Assay, 
              values_from = Calc..Concentration) #52 rows


olink_only <- distinct(olink_only) 


```

new 231008
```{r}

names(olink_only)

colnames(olink_only)[102]
colnames(olink_only)[102] <- "IL_1beta"

colnames(olink_only)[103]
colnames(olink_only)[103] <- "IL_6"

colnames(olink_only)[104]
colnames(olink_only)[104] <- "IL_8"

colnames(olink_only)[106]
colnames(olink_only)[106] <- "hu_BDNF"

colnames(olink_only)[107]
colnames(olink_only)[107] <- "hu_betaNGF"

colnames(olink_only)[108]
colnames(olink_only)[108] <- "Abeta_38"

colnames(olink_only)[109]
colnames(olink_only)[109] <- "Abeta_40"

colnames(olink_only)[110]
colnames(olink_only)[110] <- "Abeta_42"

```


export protein key
```{r}

protein_key <- names(olink_only)[c(9:100, 102:110)]

protein_key <- data.frame(protein_key)

rm(protein_key)

protein_dictionary <- data.frame(olink_metadata$protein_dictionary)

rm(protein_dictionary)

```


missing graph depiction
```{r}

library(naniar)
library(visdat)

```

visualize missing data across the olink proteins
source: https://cran.r-project.org/web/packages/visdat/vignettes/using_visdat.html
source: http://naniar.njtierney.com/articles/getting-started-w-naniar.html#exploring-missingness-relationships
```{r}

vis_dat(olink_only)

vis_miss(olink_only, sort_miss = T) + ggtitle("Missing data among OLINK proteins")

```


```{r}

missing_olink <- miss_var_summary(olink_only[, c(9:100, 102:110)])

```

exclude olink proteins with too extensive missing
exclusion criterion missing: >=50% missing
```{r}

missing_olink_cutoff <- missing_olink[missing_olink$pct_miss >= 70, ]

olink_proteins_to_exclude <- missing_olink_cutoff$variable

rm(missing_olink, missing_olink_cutoff)

```


```{r}


#olink_proteins_to_exclude <- c("BETA_NGF", "CLEC10A", "BMP_4", "NEP", "WFIKKN1", "FC_RL2", "IL_5R_ALPHA", "LAT", "CDH3", "CDH6", "NAAA",
#                               "PRTG", "GDF_8", "MDGA1", "SIGLEC1", "LXN", "SMPD1", "RSPO1")


olink_only <- olink_only[, !names(olink_only) %in% olink_proteins_to_exclude]

rm(olink_proteins_to_exclude)


```


##new 240714: combine the protein data wih age data
import data on pat age and sex
```{r}

pat_meta <- read.csv("./rev1_data/patdf_use_240714_CL.csv", header = T, sep = ";", dec = ",", na.strings = c(NA, ""))

pat_meta <- pat_meta %>%
  dplyr::select(c(patientID_use, Age, Sex))

str(pat_meta$patientID_use)
pat_meta$patientID_use <- as.character(pat_meta$patientID_use)


```

merge olink_only and pat_meta based on patientID_use
```{r}

olink_only_r1 <- full_join(pat_meta, olink_only, by = "patientID_use")

dim(olink_only_r1) 

olink_only_r1 <- distinct(olink_only_r1) #new 240715; important


```


**NEW IMPUTATIONS**
imputations 
we will use the mice package
source: https://rdrr.io/github/stefvanbuuren/mice/f/vignettes/overview.Rmd
source: https://stefvanbuuren.name/fimd/workflow.html
other interesting options could be e.g. "simputation" package
```{r}

library(mice)

imp <- mice(olink_only_r1, maxit = 10, m = 20, seed = 123, print = T)

```

qc imputations
```{r}

imp$method 

list_imputed <- list()


for(i in 1:20){
 
  imputed_data <- complete(imp, i)
  list_imputed[[paste0("imputation_", i)]] <- imputed_data
  
}


```


```{r}

rm(imp_old, imputed_data_old, list_imputed_old)

```


#DIMENSIONALITY REDUCTION TECHNIQUE
simplistic graphic depiction / tSNE
https://www.r-bloggers.com/2019/05/quick-and-easy-t-sne-analysis-in-r/
https://www.geeksforgeeks.org/how-to-make-t-sne-plot-in-r/

```{r}

library(Rtsne)

```


one imputation
```{r}

set.seed(29)

tsne1 <- Rtsne(list_imputed$imputation_1[, c(11:90, 92:100)],
                dims = 2, 
                perplexity = 17,
                check_duplicates = F, 
                pca = T, 
                is_distance = F, 
                theta = 0.0)
  
plot(tsne1$Y, col = factor(list_imputed$imputation_1$apoEgenotype), asp = 1)

tsne1_d <- as.data.frame(tsne1$Y)
  
p_tsne1_genotype1 <- ggplot(tsne1_d, aes(x = V1, y = V2, col = list_imputed$imputation_1$apoEgenotype)) +
  geom_point(size = 4.5) +
  theme_minimal() + 
  scale_color_manual(values = c("#edf8b1", "#7fcdbb", "#1d91c0", "#253494"),
                     name = "Genotype") +
  xlab("Dimension 1") + ylab("Dimension 2") + 
  ggtitle("tSNE of Olink proteins", subtitle = "depending on SNP status, n = 1 imputation")


```

```{r}

rm(tsne1, tsne1_d, p_tsne1_genotype1)

```

generalized solution: for loop
```{r}

tsne_data <- list()

for(i in 1:20){
  
  set.seed(29)
  
  interim <- Rtsne(list_imputed[[i]][, c(11:90, 92:100)],
                dims = 2, 
                perplexity = 17,
                check_duplicates = F, 
                pca = T, 
                is_distance = F, 
                theta = 0.0)
  
  interim_d <- as.data.frame(interim$Y)
  
  tsne_data[[paste0("tsne_imputation_", i)]] <- interim_d
  
  
}


d_tsne_mean <- Reduce("+", tsne_data)

d_tsne_mean <- d_tsne_mean/length(tsne_data)


```


visualize
```{r}

p_tsne_time <- ggplot(d_tsne_mean, aes(x = V1, y = V2,
                             col = list_imputed$imputation_1$sample_time)) +
  geom_point(size = 4.5) +
  theme_minimal() +
  scale_color_manual(values = c("#bfd3e6", "#8c6bb1"), labels = c("sample1", "sample2"), 
                     name = "sampling time") +
  xlab("Dimension 1") + ylab("Dimension 2") +
  ggtitle("tSNE of Olink proteins", subtitle = "depending on sampling time, n = 20 imputations")


p_tsne_genotype <- ggplot(d_tsne_mean, aes(x = V1, y = V2, col = list_imputed$imputation_1$apoEgenotype)) +
  geom_point(size = 4.5) +
  theme_minimal() + 
  scale_color_manual(values = c("#edf8b1", "#7fcdbb", "#1d91c0", "#253494"),
                     name = "Genotype") +
  xlab("Dimension 1") + ylab("Dimension 2") + 
  ggtitle("tSNE of Olink proteins", subtitle = "depending on SNP status, n = 20 imputations")


p_tsne_genotype_time <- ggplot(d_tsne_mean, aes(x = V1, y = V2, col = list_imputed$imputation_1$apoEgenotype)) +
  geom_point(aes(shape = list_imputed$imputation_1$sample_time), size = 4.5) +
  theme_minimal() + 
  scale_color_manual(values = c("#edf8b1", "#7fcdbb", "#1d91c0", "#253494"),
                     name = "Genotype") +
  xlab("Dimension 1") + ylab("Dimension 2") + 
  ggtitle("tSNE of Olink proteins longitudinally", subtitle = "depending on SNP status, n = 20 imputations")



```


```{r}

rm(p_tsne_time, p_tsne_genotype, p_tsne_genotype_time)

rm(tsne_data, interim, imputed_data, interim_d, d_tsne_mean, i, vars_to_remove)

```

#CLUSTERING

**clustering analysis**
```{r}

library(ComplexHeatmap)
library(RColorBrewer)
library(circlize)


```


determine number of clusters (one imputation)
```{r}

#select and scale full data

cluster_d <- list_imputed$imputation_1

rownames(cluster_d) <- paste(cluster_d$Tsample_id, cluster_d$Olink_patientID) 

cluster_d <- cluster_d[, c(11:90, 92:100)]

colnames(cluster_d) <- names(cluster_d)

cluster_d_scale <- scale(cluster_d)


```

determine number of clusters (one imputation) cont
```{r}

#distance calculation

library(cluster)
library(factoextra)

fviz_dist(get_dist(cluster_d, method = "euclidean", stand = T))
fviz_dist(get_dist(cluster_d, method = "spearman", stand = T))
fviz_dist(get_dist(cluster_d, method = "kendall", stand = T))


```

determine number of clusters (one imputation) cont
```{r}

#assess clustering tendency

#use scaled data

fviz_pca_ind(prcomp(scale(cluster_d_scale)), habillage = list_imputed$imputation_1$apoEgenotype, geom = "point", ggtheme = theme_classic()) #no obvious

fviz_pca_ind(prcomp(scale(cluster_d_scale)), habillage = list_imputed$imputation_1$sample_time, geom = "point", ggtheme = theme_classic()) #seems as if there might be clustering


```

determine number of clusters (one imputation) cont
https://www.r-bloggers.com/assessing-clustering-tendency-a-vital-issue-unsupervised-machine-learning/
https://www.datanovia.com/en/lessons/assessing-clustering-tendency/ <-- here it says that the H threshold is 0.5
```{r}

#statistical methods to assess clustering tendency

get_clust_tendency(cluster_d_scale, n = nrow(cluster_d_scale) - 1, graph = T) #$hopkins_stat (H): 0.642985 (a little bit lower than the original code)

#maybe worth trying clustering techniques 

```

determine number of clusters (one imputation) cont
```{r}

library(NbClust)


NbClust(cluster_d_scale, distance = "euclidean",
        min.nc = 2, max.nc = 10, method = "kmeans",
        index = "tau")

#frey: 1
#kl, ch, duda, pseudot2, beale, gap, mcclain: 2
#ptbiserial, tau, ball, hubert(?): 3
#cindex: 4
#ratkowsky: 5
#hartigan, silhouette: 7
#sdindex, dunn: 9
#db, sdbw, gamma, gplus: 10
#dindex: unclear

#ccc, scott, marriot, trcovw, tracew, friedman, rubin does not work


#in total n = 30 different clustering algorithms
#of these, n = 23 works

# of these n = 7 (majority) suggest n = 2 clusters
# of these n = 4 (second most common) suggest n = 3 or n = 10 clusters

```

validate clustering strategy
```{r}

set.seed(29)

partCl_2 <- kmeans(cluster_d_scale, 2, nstart = 25)
fviz_cluster(partCl_2, data = cluster_d_scale, ellipse.type = "convex", ggtheme = theme_minimal())
#here - the two sampling time points for each patients tend to fall into the same cluster - time might be a meaningful clustering variable

partCl_4 <- kmeans(cluster_d_scale, 3, nstart = 25)
fviz_cluster(partCl_4, data = cluster_d_scale, ellipse.type = "convex", ggtheme = theme_minimal())
#240715: here n = 3 clusters seems possibly more meaningful 

```

validate clustering strategy cont
```{r}

library(clValid)

clmethods <- c("hierarchical", "kmeans", "pam")
intern <- clValid(cluster_d_scale, nClust = 2:6,
                  clMethods = clmethods,
                  validation = "internal")

summary(intern) 
#suggested strategy: hierarchical clustering, n = 2 clusters


```

depict hierarchical clustering
```{r}

#full data
#matrix format + scaled

cluster_d_scale[1:5, 1:5]

dist_h1 <- dist(cluster_d_scale, method = "euclidean")

hierCl <- hclust(d = dist_h1, method = "ward.D2")

fviz_dend(hierCl, cex = .5)

f_coph <- cophenetic(hierCl)

cor(dist_h1, f_coph) #0.4413906 (updated)


hierCl_2 <- hclust(d = dist_h1, method = "average")
cor(dist_h1, cophenetic(hierCl_2)) #0.7795342

hierCl_3 <- hclust(d = dist_h1, method = "complete")
cor(dist_h1, cophenetic(hierCl_3)) #0.5899953


#linkage method should be "average" based on above calc

fviz_dend(hierCl_2, cex = 0.5) 

fviz_dend(hierCl_3, cex = 0.5) 


f_agnes <- agnes(x = cluster_d_scale,
                 stand = F,
                 metric = "euclidean", 
                 method = "ward")

fviz_dend(f_agnes, cex = 0.5)

f_diana <- diana(x = cluster_d_scale,
                 stand = F,
                 metric = "euclidean")

fviz_dend(f_diana, cex = 0.5)

#agnes or hierCl look best graphically

```

```{r}

rm(cluster_d, cluster_d_scale,
   f_agnes, f_diana, hierCl, hierCl_2, hierCl_3,
   partCl_2, partCl_4, clmethods,
   intern,
   dist_h1, f_coph)

```

generalize cluster determination for all imputations
```{r}

cluster_d <- list()
cluster_d_scale <- list()

for(i in 1:20){
  
  interim <- list_imputed[[i]]
  
  rownames(interim) <- paste(interim$Tsample_id, interim$Olink_patientID)
  
  interim <- interim[, c(11:90, 92:100)]
  
  colnames(interim) <- names(interim)
  
  cluster_d[[paste0("imputation_", i)]] <- interim
  
  interim_scale <- scale(interim)
  
  cluster_d_scale[[paste0("imputation_", i)]] <- interim_scale
  
}

```

for the generalized solution we prioritize among the different analyses
clustering analyses that are meaningful to do: 
1. calculation of hopkins statistic 

calculation of hopkins statistic
```{r}

hopkins <- list()

for(i in 1:20){
  
  interim <- get_clust_tendency(cluster_d_scale[[i]], n = nrow(cluster_d_scale[[i]]) - 1, graph = T)
  
  hopkins[[paste0("imputation_", i)]] <- interim$hopkins_stat
  
}


hopkins_mean <- Reduce("+", hopkins)

hopkins_mean <- hopkins_mean/length(hopkins)

hopkins_mean #[1] 0.6493411 (2024-07-15)

#maybe worth trying clustering techniques 

```

```{r}

rm(hopkins, interim, hopkins_mean, i)

```

determine number of clusters generalized utilizing NbClust
i) run clustering algorithm across imputations
j) run clustering algorithm across clustering methods 
```{r}

set.seed(29)

cluster_indices <- c("kl", "ratkowsky", "ptbiserial", "tau", "ch", "duda", "pseudot2", "beale", "gap", "mcclain", "hartigan", 
                     "cindex", "db", "sdindex", "sdbw", "silhouette", "ball", "frey", "gamma", "gplus", "dunn")

cluster_count <- list()

for(i in names(cluster_d_scale)[1:20]){
  
  for(j in cluster_indices[1:21]){
    
    
    interim2 <- NbClust(cluster_d_scale[[i]], distance = "euclidean",
        min.nc = 2, max.nc = 10, method = "kmeans",
        index = j)
  
  
    cluster_count[[j]][[i]] <- interim2$Best.nc[1]
    
    
  } 
  
  #this syntax works for:
  #kl, ratkowsky, ptbiserial, tau, ch, duda, pseudot2, beale, gap, mcclain, hartigan
  #cindex, db, sdindex, sdbw, silhouette, ball, frey, gamma, gplus, dunn
  
  #this syntax does NOT work for:
  #hubert, dindex
  
  #the clustering algorithms don't work at all: 
  ##ccc, scott, marriot, trcovw, tracew, friedman, rubin does not work

  
  
}



```

```{r}

cluster_count_df <- data.frame(unlist(cluster_count))

cluster_count_df$meta <- rownames((cluster_count_df))

cluster_count_df <- cluster_count_df %>%
  separate(meta, into = c("clusterMethod", "imputation_index", "variable"), sep = "\\.")

names(cluster_count_df)

names(cluster_count_df)[1] <- "value"


```

calculate the suggested numbers of clusters from each clustering method
```{r}

str(cluster_count_df) #value is numerical

cluster_count_df_sum <- cluster_count_df %>%
  group_by(clusterMethod) %>%
  summarise(mean_no_of_clusters = mean(value, na.rm = T),
            median_no_of_clusters = median(value, na.rm = T)
            )

summary(cluster_count_df_sum)

#21 clusters

table(cluster_count_df_sum$mean_no_of_clusters)

table(cluster_count_df_sum$median_no_of_clusters) #2 clusters: 7 instances (majority). still holds 240715

```

it thus seems likely that we should structure the data into n = 2 clusters if anything
```{r}

rm(cluster_count_df_sum, cluster_count_df, cluster_count,
   cluster_indices, i, j, interim2)


```

now we have decided the number of clusters (n = 2) - now it makes sense to decide clustering algorithm (hierarchical, kmeans, pam)
```{r}

library(clValid)
library(mclust)

clmethods <- c("hierarchical", "kmeans", "diana", "fanny", "model", "sota", "pam", "clara", "agnes")

cl_validation <- list()

for(i in names(cluster_d_scale)[1:20]){

  set.seed(29)
  
  interim <- clValid(cluster_d_scale[[i]], nClust = 2:6,
                  clMethods = clmethods,
                  validation = "internal")

  cl_validation[[i]] <- interim
  
  
}


```

```{r}

summary(cl_validation$imputation_20)


#all: hier; n = 2 clusters. 

#evaluated with regard to connectivity, dunn, silhouette score


```

```{r}

rm(clmethods, interim, cl_validation, i)

```

graphical depiction clustering
matrix format (scaled imputations)
```{r}

distance <- list()

for(i in names(cluster_d_scale)[1:20]){
  
  interim <- dist(cluster_d_scale[[i]], method = "euclidean")
  
  interim2 <- hclust(d = interim, method = "ward.D2")
  
  distance[[i]] <- interim
  
  
}



```

```{r}

distance$imputation_1 #seems as if col and row names are identical

distance$imputation_12 #seems as if col and row names are identical

distance$imputation_19 #seems as if col and row names are identical

distance_mean <- Reduce("+", distance)

distance_mean

distance_mean <- distance_mean/length(distance)

distance_mean

```

```{r}

hier_cluster <- hclust(d = distance_mean, method = "ward.D2")

fviz_dend(hier_cluster, cex = .5)

f_coph <- cophenetic(hier_cluster)

cor(distance_mean, f_coph) 


#new corr 240715 (ward.D2): [1] 0.4677026


```

```{r}

rm(distance, distance_mean, hier_cluster, f_coph, interim, interim2, i)

```

#HEATMAP
visualize heatmap (one imputation)
```{r}

set.seed(29)

#pending clustering
hierCl_use <- hcut(cluster_d$imputation_1, k = 2, hc_func = "hclust", hc_method = "ward.D2", hc_metric = "euclidean", stand = T) 


#add categories
col <- list(genotype = c("e4/e3" = "#1d91c0", "e3/e3" = "#7fcdbb", "e3/e2" = "#edf8b1", "e4/e4" = "#253494"),
           sample_time = c("d1" = "#bfd3e6", "d3" = "#8c6bb1"))


#individuals as rows
ra <- rowAnnotation(
  genotype = list_imputed$imputation_1$apoEgenotype,
  sample_time = list_imputed$imputation_1$sample_time,
  patient_ID = list_imputed$imputation_1$Olink_patientID,
  col = col
)


hm_imp1 <- Heatmap(scale(list_imputed$imputation_1[, c(11:90, 92:100)]),
                   name = "scaled protein levels",
                   column_title = "Proteins",
                   column_names_gp = gpar(fontsize = 5),
                   show_column_dend = T,
                   show_row_dend =  T,
                   right_annotation = ra
                   #, split = hierCl_use$cluster
                   )


draw(hm_imp1) ; row_order(hm_imp1)

```

source: https://rdrr.io/bioc/ComplexHeatmap/man/row_order-Heatmap-method.html

verify that individual patient rows can be identified accurately
```{r}

row_indices_test <- as.numeric(as.character(row_order(hm_imp1))) #core function row_order

row_indices_id <- list_imputed$imputation_1$Olink_patientID[row_indices_test]

test_matrix <- hm_imp1@matrix

class(test_matrix)

test_test_matrix <- as.data.frame(test_matrix)

test_test_matrix$Olink_patientID <- row_indices_id

```

remove redundant code
```{r}

rm(cluster_d, cluster_d_scale, col, hierCl_use, hm_imp1, interim_scale, ra, test_matrix, test_test_matrix, 
   row_indices_id, row_indices_test)

```

##access the underlying data: hm_imp1@matrix --> maybe pool as for tsne?
##generalize and pool heatmaps
```{r}

hm_imp_p <- list()
hm_imp_d <- list()
row_number <- list()


for(i in 1:20){
  
  set.seed(29)
  
  #add categories
  col <- list(genotype = c("e4/e3" = "#1d91c0", "e3/e3" = "#7fcdbb", "e3/e2" = "#edf8b1", "e4/e4" = "#253494"),
           sample_time = c("d1" = "#bfd3e6", "d3" = "#8c6bb1"))


  #individuals as rows
  ra <- rowAnnotation(
    genotype = list_imputed[[i]]$apoEgenotype,
    sample_time = list_imputed[[i]]$sample_time,
    patient_ID = list_imputed[[i]]$Olink_patientID,
    col = col
  )

  
  interim <- Heatmap(scale(list_imputed[[i]][, c(11:90, 92:100)]),
                   name = "scaled OLINK levels",
                   column_title = "OLINK proteins",
                   column_names_gp = gpar(fontsize = 5),
                   show_column_dend = T,
                   show_row_dend =  T,
                   right_annotation = ra
                   )
  
  interim_d <- interim@matrix
  
  hm_imp_p[[paste0("heatmap_imputation_", i)]] <- interim
  
  hm_imp_d[[paste0("heatmap_imputation_", i)]] <- interim_d 
  
  row_number[[paste0("heatmap_imputation_", i)]] <- row_order(hm_imp_p[[paste0("heatmap_imputation_", i)]])
  
}



```

add Olink_patientID
```{r}

str(row_number$heatmap_imputation_1) #int

hm_imp_df <- list()


for(i in 1:20){
  
  interim <- list_imputed[[i]]$Olink_patientID[row_number[[i]]]
  
  
  interim2 <- as.data.frame(hm_imp_d[[i]]) 
  
  interim2$Olink_patientID <- interim
  
  interim2$imputation_index <- rep(i, length.out = dim(interim2)[1])
  
  
  hm_imp_df[[paste0("heatmap_imputation_", i)]] <- interim2
  
  
}



```

```{r}


hm_imp_df_all <- bind_rows(hm_imp_df)

```

calculate mean
```{r}

hm_imp_df_all_sum <- hm_imp_df_all %>%
  group_by(Olink_patientID) %>%
  summarise(across(NMNAT1:Abeta_42, \(x) mean(x, na.rm = T)))


```

merge imputed and pooled heatmap data with the other metadata
```{r}

meta_vars <- names(olink_only_r1)[c(1:10, 91)]

olink_only_r1_subset <- olink_only_r1[, meta_vars]

hm_imp_df_all_sum_meta <- full_join(hm_imp_df_all_sum, olink_only_r1_subset, by = "Olink_patientID")

```

```{r}

set.seed(29)
  
  #add categories
  col <- list(genotype = c("e4/e3" = "#1d91c0", "e3/e3" = "#7fcdbb", "e3/e2" = "#edf8b1", "e4/e4" = "#253494"),
           sample_time = c("d1" = "#bfd3e6", "d3" = "#8c6bb1"))


  #individuals as rows
  ra <- rowAnnotation(
    genotype = hm_imp_df_all_sum_meta$apoEgenotype,
    sample_time = hm_imp_df_all_sum_meta$sample_time,
    patient_ID = hm_imp_df_all_sum_meta$Olink_patientID,
    col = col
  )


hm_pooled <- Heatmap(scale(hm_imp_df_all_sum_meta[, c(2:90)]),
                   name = "scaled, imputed OLINK levels",
                   column_title = "OLINK proteins",
                   column_names_gp = gpar(fontsize = 5),
                   show_column_dend = T,
                   show_row_dend =  T,
                   right_annotation = ra
                   )

draw(hm_pooled)


```

```{r}

rm(col, hm_imp_d, hm_imp_df, hm_imp_df_all, hm_imp_df_all_sum, hm_imp_df_all_sum_meta, hm_imp_p, hm_pooled,
   interim_d, interim2, i, interim, meta_vars, row_number, ra)

```


#REGRESSION ANALYSES
in older version of this script regular linear regression analyses were included
this has been omitted now as iid criteria is not met

this code chunk has been moved to the top of this section
```{r}

##flatten list code
#source: https://stackoverflow.com/questions/55731855/how-to-extract-every-ggplot2-plot-from-a-nested-list
FlattenListofGgplot <- function(x){
  morelists <- sapply(x, function(xprime){
    "list" %in% class(xprime) & !("gg" %in% class(xprime))
  })
  
  out <- c(x[!morelists], unlist(x[morelists], recursive = F))
  
  if(sum(morelists)){
    Recall(out)
  }
  else{
    return(out)
  }
  
}



```


prep data - new 231031 starts
Ratio Abeta42/40 regression to apoe/time
```{r}

list_imputed_ratio <- list()

for(i in names(list_imputed)[1:20]){
  
  interim <- list_imputed[[i]] %>%
    group_by(Tsample_id, Olink_patientID) %>%
    mutate(Abeta42_40_ratio = Abeta_42/Abeta_40)
  
  list_imputed_ratio[[i]] <- interim
  
}



```


prep data
```{r}

list_imputed_long <- list()


for(i in names(list_imputed_ratio)[1:20]){
  
  interim <- list_imputed_ratio[[i]] %>%
    group_by(Tsample_id, Olink_patientID) %>%
    pivot_longer(cols = c(NMNAT1:KYNU, IL_1beta:Abeta42_40_ratio), 
               names_to = "olink_protein",
               values_to = "value")
  
  list_imputed_long[[i]] <- interim
  
  
}



```

```{r}

rm(list_imputed_ratio)

```

prep data
```{r}

#create var "apoE_e4_carrier"

for(i in names(list_imputed_long)[1:20]){
  
  interim <- if_else(grepl("e4", list_imputed_long[[i]]$apoEgenotype), 1, 0)
  
  list_imputed_long[[i]]$apoE_e4_carrier <- interim
  
}


```

prep data
```{r}

olink_protein_names <- unique(list_imputed_long$imputation_1$olink_protein)

```

MIXED MODELS


```{r}
library(nlme)

```


MIXED MODELS (UNIVARIATE: AGE)
```{r}

age_univar_mimo <- list()
age_univar_mimo_p <- list()
age_univar_mimo_coef <- list()


for(i in names(list_imputed_long)[1:20]){
  
  for(j in olink_protein_names[1:90]){
    
    interim <- list_imputed_long[[i]] %>%
      filter(olink_protein == j)
    
    interim2 <- summary(lme(value ~ Age, 
                    random = ~1 | patientID_use,
                    data = interim))
    
    age_univar_mimo[[j]][[i]] <- interim2
    
    age_univar_mimo_p[[j]][[i]] <- interim2$tTable[10]
    
    age_univar_mimo_coef[[j]][[i]] <- interim2$tTable[2]
  
  }
  
  
}



```

```{r}

age_univar_mimo_df <- data.frame(unlist(age_univar_mimo_p))
age_univar_mimo_coef_df <- data.frame(unlist(age_univar_mimo_coef))

table(age_univar_mimo_df$unlist.age_univar_mimo_p. <= 0.05) #T: 84 

```

generalized solution reg assumptions
```{r}

model_out <- list()
original_out <- list()
p_vari <- list()
p_qq <- list()
p_corr <- list()
p_pacf <- list()


for(i in names(list_imputed_long)[1:20]){
  
  for(j in olink_protein_names[1:90]){

        original_data <- list_imputed_long[[i]] %>%
          filter(olink_protein == j)
        
        
        model <- lme(value ~ Age, 
                    random = ~1 | patientID_use,
                    data = original_data)
    
  
        original_data$fitted <- predict(model)
        
        original_data$resid <- residuals(model)
        
        
        
        ##residual vs fitted
        p1 <- ggplot(original_data, aes(fitted, resid)) + 
              geom_point() +
              geom_hline(yintercept = 0) +
              stat_smooth(method = "loess") +
              theme_minimal() +
              ggtitle(paste(j, i))
        
        p_vari[[j]][[i]] <- p1
        
        
        
        ##normality plot for the residuals
        p2 <- ggplot(original_data, aes(sample = resid)) +
              stat_qq() + stat_qq_line() +
              theme_minimal() +
              ggtitle(paste(j, i))
        
        p_qq[[j]][[i]] <- p2
        
  
        
        ##corr fitted vs observed
        p3 <- ggplot(original_data, aes(fitted, value)) +
              geom_point() +
              stat_smooth(method = "loess") + 
              theme_minimal() +
              ggtitle(paste(j, i))
        
        p_corr[[j]][[i]] <- p3
        
        library(forecast)

        p4 <- ggtsdisplay(original_data$resid)
        
        p_pacf[[j]][[i]] <- p4
        
  
    
  }
  
  
  

}



```


save all files
```{r}

library(gridExtra)

p_vari_s <- FlattenListofGgplot(p_vari)

p_qq_s <- FlattenListofGgplot(p_qq)

p_corr_s <- FlattenListofGgplot(p_corr)

p_pacf_s <- FlattenListofGgplot(p_pacf)


```


```{r}

rm(p1, p2, p3, p4, p_vari, p_vari_s, p_qq, p_qq_s, p_corr, p_corr_s, p_pacf, p_pacf_s, original_out, model_out,
   original_data, model)

```


```{r}

age_univar_mimo_coef_df$meta <- rownames(age_univar_mimo_coef_df)

age_univar_mimo_df$meta <- rownames(age_univar_mimo_df)

age_univar_mimo_df <- full_join(age_univar_mimo_df, age_univar_mimo_coef_df, by = "meta")


age_univar_mimo_df <- age_univar_mimo_df %>%
  separate(meta, into = c("olink_protein", "imputation_index"), sep = "\\.")

head(age_univar_mimo_df)

names(age_univar_mimo_df)[1] <- "p_age_unadj"
names(age_univar_mimo_df)[4] <- "coef_age"

age_univar_mimo_df <- age_univar_mimo_df %>%
  group_by(olink_protein) %>%
  summarise(across(c(p_age_unadj, coef_age), list(mean = ~ mean(.x, na.rm = T))))
    
age_univar_mimo_df$p_age_adj <- p.adjust(age_univar_mimo_df$p_age_unadj_mean, method = "BH")

age_univar_mimo_df_sign <- age_univar_mimo_df %>%
  filter(p_age_unadj_mean <= 0.05) 

View(age_univar_mimo_df_sign) 

age_univar_mimo_df_sign <- age_univar_mimo_df_sign[, c(1, 3, 2, 4)]


```


```{r}

rm(age_univar_mimo, age_univar_mimo_df_sign, age_univar_mimo, age_univar_mimo_p)

```

ADDITIVE ANALYSIS: AGE
```{r}

library(nlme)

rev1_df_age <- list()
rev1_p_age <- list()

for(i in names(list_imputed_long)[1:20]){
  
  for(j in olink_protein_names[1:90]){
    
    interim <- list_imputed_long[[i]] %>%
      filter(olink_protein == j)
    
    interim2 <- summary(lme(value ~ apoE_e4_carrier + Age, 
                    random = ~1 | patientID_use,
                    data = interim))
    
    rev1_df_age[[j]][[i]] <- interim2
    
    rev1_p_age[[j]][[i]] <- interim2$tTable[15]
    
    
  }
  
  
}



```

POOL AGE ANALYSIS
```{r}

rev1_p_age_df <- data.frame(unlist(rev1_p_age))

rev1_p_age_df$meta <- rownames(rev1_p_age_df)

rev1_p_age_df <- rev1_p_age_df %>%
  separate(meta, into = c("olink_protein", "imputation_index"), sep = "\\.")

names(rev1_p_age_df)[1] <- "p_rev1_age"

rev1_p_age_df <- rev1_p_age_df %>%
  group_by(olink_protein) %>%
  summarise(p_rev1_age_unadj_mean = mean(p_rev1_age, na.rm = T))

rev1_p_age_df$p_rev1_age_adj <- p.adjust(rev1_p_age_df$p_rev1_age_unadj_mean, method = "BH")


```

```{r}

rm(rev1_p_age_df, rev1_p_age, rev1_p_age_df, rev1_df_age)

```


```{r}

model_out <- list()
original_out <- list()
p_vari <- list()
p_qq <- list()
p_corr <- list()
p_pacf <- list()


for(i in names(list_imputed_long)[1:20]){
  
  for(j in olink_protein_names[1:90]){

        original_data <- list_imputed_long[[i]] %>%
          filter(olink_protein == j)
        
        model <- lme(value ~ apoE_e4_carrier + Age, 
                    random = ~1 | patientID_use,
                    data = original_data)

  
        original_data$fitted <- predict(model)
        
        original_data$resid <- residuals(model)
        

        
        ##residual vs fitted
        p1 <- ggplot(original_data, aes(fitted, resid)) + 
              geom_point() +
              geom_hline(yintercept = 0) +
              stat_smooth(method = "loess") +
              theme_minimal() +
              ggtitle(paste(j, i))
        
        p_vari[[j]][[i]] <- p1
        
        
        
        ##normality plot for the residuals
        p2 <- ggplot(original_data, aes(sample = resid)) +
              stat_qq() + stat_qq_line() +
              theme_minimal() +
              ggtitle(paste(j, i))
        
        p_qq[[j]][[i]] <- p2
        
          
        
        ##corr fitted vs observed
        p3 <- ggplot(original_data, aes(fitted, value)) +
              geom_point() +
              stat_smooth(method = "loess") + 
              theme_minimal() +
              ggtitle(paste(j, i))
        
        p_corr[[j]][[i]] <- p3
        
        library(forecast)

        p4 <- ggtsdisplay(original_data$resid)
        
        p_pacf[[j]][[i]] <- p4
        
  
    
  }
  
  
  

}



```


save all files
```{r}

library(gridExtra)

p_vari_s <- FlattenListofGgplot(p_vari)

p_qq_s <- FlattenListofGgplot(p_qq)

p_corr_s <- FlattenListofGgplot(p_corr)

p_pacf_s <- FlattenListofGgplot(p_pacf)



```

```{r}

rm(p1, p2, p3, p4, p_vari, p_vari_s, p_qq, p_qq_s, p_corr, p_corr_s, p_pacf, p_pacf_s, original_out, model_out,
   original_data, model)

```

MIXED MODELS (UNIVARIATE: SEX)
```{r}

sex_univar_mimo <- list()
sex_univar_mimo_p <- list()


for(i in names(list_imputed_long)[1:20]){
  
  for(j in olink_protein_names[1:90]){
    
    interim <- list_imputed_long[[i]] %>%
      filter(olink_protein == j)
    
    interim2 <- summary(lme(value ~ Sex, 
                    random = ~1 | patientID_use,
                    data = interim))
    
    sex_univar_mimo[[j]][[i]] <- interim2
    
    sex_univar_mimo_p[[j]][[i]] <- interim2$tTable[10]
  
  }
  
  
}



```

```{r}

sex_univar_mimo_df <- data.frame(unlist(sex_univar_mimo_p))

table(sex_univar_mimo_df$unlist.sex_univar_mimo_p. <= 0.05) #T: 5 

```


```{r}

model_out <- list()
original_out <- list()
p_vari <- list()
p_qq <- list()
p_corr <- list()
p_pacf <- list()


for(i in names(list_imputed_long)[1:20]){
  
  for(j in olink_protein_names[1:90]){

        original_data <- list_imputed_long[[i]] %>%
          filter(olink_protein == j)
        
        model <- lme(value ~ Sex, 
                    random = ~1 | patientID_use,
                    data = original_data)
    
  
        original_data$fitted <- predict(model)
        
        original_data$resid <- residuals(model)
        

        
        ##residual vs fitted
        p1 <- ggplot(original_data, aes(fitted, resid)) + 
              geom_point() +
              geom_hline(yintercept = 0) +
              stat_smooth(method = "loess") +
              theme_minimal() +
              ggtitle(paste(j, i))
        
        p_vari[[j]][[i]] <- p1
        
        
        
        ##normality plot for the residuals
        p2 <- ggplot(original_data, aes(sample = resid)) +
              stat_qq() + stat_qq_line() +
              theme_minimal() +
              ggtitle(paste(j, i))
        
        p_qq[[j]][[i]] <- p2
        
  
        
        ##corr fitted vs observed
        p3 <- ggplot(original_data, aes(fitted, value)) +
              geom_point() +
              stat_smooth(method = "loess") + 
              theme_minimal() +
              ggtitle(paste(j, i))
        
        p_corr[[j]][[i]] <- p3
        
      
        library(forecast)

        p4 <- ggtsdisplay(original_data$resid)
        
        p_pacf[[j]][[i]] <- p4
        
  
    
  }
  
  
  

}



```


save all files
```{r}

library(gridExtra)

p_vari_s <- FlattenListofGgplot(p_vari)

p_qq_s <- FlattenListofGgplot(p_qq)

p_corr_s <- FlattenListofGgplot(p_corr)

p_pacf_s <- FlattenListofGgplot(p_pacf)



```


```{r}

rm(p1, p2, p3, p4, p_vari, p_vari_s, p_qq, p_qq_s, p_corr, p_corr_s, p_pacf, p_pacf_s, original_out, model_out,
   original_data, model)

```


```{r}

sex_univar_mimo_df$meta <- rownames(sex_univar_mimo_df)

sex_univar_mimo_df <- sex_univar_mimo_df %>%
  separate(meta, into = c("olink_protein", "imputation_index"), sep = "\\.")

head(sex_univar_mimo_df)

names(sex_univar_mimo_df)[1] <- "p_univar_unadj"

sex_univar_mimo_df <- sex_univar_mimo_df %>%
  group_by(olink_protein) %>%
  summarise(p_univar_unadj_mean = mean(p_univar_unadj, na.rm = T))

sex_univar_mimo_df$p_univar_adj <- p.adjust(sex_univar_mimo_df$p_univar_unadj_mean, method = "BH")

sex_univar_mimo_df_sign <- sex_univar_mimo_df %>%
  filter(p_univar_unadj_mean <= 0.05) 

sex_univar_mimo_df_sign #empty


```

```{r}

rm(sex_univar_mimo, sex_univar_mimo_df_sign, sex_univar_mimo, sex_univar_mimo_p)

```


ADDITIVE ANALYSIS: SEX
```{r}

library(nlme)

rev1_df_sex <- list()
rev1_p_sex <- list()

for(i in names(list_imputed_long)[1:20]){
  
  for(j in olink_protein_names[1:90]){
    
    interim <- list_imputed_long[[i]] %>%
      filter(olink_protein == j)
    
    interim2 <- summary(lme(value ~ apoE_e4_carrier + Sex, 
                    random = ~1 | patientID_use,
                    data = interim))
    
    rev1_df_sex[[j]][[i]] <- interim2
    
    rev1_p_sex[[j]][[i]] <- interim2$tTable[15]
    
    
  }
  
  
}



```

POOL SEX ANALYSIS
```{r}

rev1_p_sex_df <- data.frame(unlist(rev1_p_sex))

rev1_p_sex_df$meta <- rownames(rev1_p_sex_df)

rev1_p_sex_df <- rev1_p_sex_df %>%
  separate(meta, into = c("olink_protein", "imputation_index"), sep = "\\.")

names(rev1_p_sex_df)[1] <- "p_rev1_sex"

rev1_p_sex_df <- rev1_p_sex_df %>%
  group_by(olink_protein) %>%
  summarise(p_rev1_sex_unadj_mean = mean(p_rev1_sex, na.rm = T))

rev1_p_sex_df$p_rev1_sex_adj <- p.adjust(rev1_p_sex_df$p_rev1_sex_unadj_mean, method = "BH")


```


```{r}

rm(rev1_p_sex_df, rev1_p_sex, rev1_p_sex_df, rev1_df_sex)

```


generalized solution - regression assumptions
```{r}

model_out <- list()
original_out <- list()
p_vari <- list()
p_qq <- list()
p_corr <- list()
p_pacf <- list()


for(i in names(list_imputed_long)[1:20]){
  
  for(j in olink_protein_names[1:90]){

        original_data <- list_imputed_long[[i]] %>%
          filter(olink_protein == j)
        
        model <- lme(value ~ apoE_e4_carrier + Sex, 
                    random = ~1 | patientID_use,
                    data = original_data)

  
        original_data$fitted <- predict(model)
        
        original_data$resid <- residuals(model)
        
        
        
        #original_out[[j]][[i]] <- original_data
        #model_out[[j]][[i]] <- model
        
        
        ##residual vs fitted
        p1 <- ggplot(original_data, aes(fitted, resid)) + 
              geom_point() +
              geom_hline(yintercept = 0) +
              stat_smooth(method = "loess") +
              theme_minimal() +
              ggtitle(paste(j, i))
        
        p_vari[[j]][[i]] <- p1
        
        
        
        ##normality plot for the residuals
        p2 <- ggplot(original_data, aes(sample = resid)) +
              stat_qq() + stat_qq_line() +
              theme_minimal() +
              ggtitle(paste(j, i))
        
        p_qq[[j]][[i]] <- p2
        
          
        
        ##corr fitted vs observed
        p3 <- ggplot(original_data, aes(fitted, value)) +
              geom_point() +
              stat_smooth(method = "loess") + 
              theme_minimal() +
              ggtitle(paste(j, i))
        
        p_corr[[j]][[i]] <- p3
        
        
        library(forecast)

        p4 <- ggtsdisplay(original_data$resid)
        
        p_pacf[[j]][[i]] <- p4
        
  
    
  }
  
  
  

}



```


save all files
```{r}

library(gridExtra)

p_vari_s <- FlattenListofGgplot(p_vari)

p_qq_s <- FlattenListofGgplot(p_qq)

p_corr_s <- FlattenListofGgplot(p_corr)

p_pacf_s <- FlattenListofGgplot(p_pacf)



```

```{r}

rm(p1, p2, p3, p4, p_vari, p_vari_s, p_qq, p_qq_s, p_corr, p_corr_s, p_pacf, p_pacf_s, original_out, model_out,
   original_data, model)

```


MIXED MODELS (INTERACTION)
```{r}

library(nlme)

univar_mimo <- list()

univar_mimo_coef_apoE_e4_carrier <- list()
univar_mimo_p_apoE_e4_carrier <- list()

univar_mimo_coef_time <- list()
univar_mimo_p_sample_time <- list()

univar_mimo_coef_interaction <- list()
univar_mimo_p_interaction <- list()


for(i in names(list_imputed_long)[1:20]){
  
  for(j in olink_protein_names[1:90]){
    
    interim <- list_imputed_long[[i]] %>%
      filter(olink_protein == j)
    
    interim2 <- summary(lme(value ~ apoE_e4_carrier * sample_time, 
                    random = ~1 | patientID_use,
                    data = interim))
    
    univar_mimo[[j]][[i]] <- interim2
    
    univar_mimo_coef_apoE_e4_carrier[[j]][[i]] <- interim2$tTable[2]
    univar_mimo_p_apoE_e4_carrier[[j]][[i]] <- interim2$tTable[18]
    
    univar_mimo_coef_time[[j]][[i]] <- interim2$tTable[3]
    univar_mimo_p_sample_time[[j]][[i]] <- interim2$tTable[19]
    
    univar_mimo_coef_interaction[[j]][[i]] <- interim2$tTable[4]
    univar_mimo_p_interaction[[j]][[i]] <- interim2$tTable[20]
    
  
  }
  
  
}


```


```{r}

model_out <- list()
original_out <- list()
p_vari <- list()
p_qq <- list()
p_corr <- list()
p_pacf <- list()


for(i in names(list_imputed_long)[1:20]){
  
  for(j in olink_protein_names[1:90]){

        original_data <- list_imputed_long[[i]] %>%
          filter(olink_protein == j)
        
          
        model <- lme(value ~ apoE_e4_carrier * sample_time, 
                    random = ~1 | patientID_use,
                    data = original_data)
    

  
        original_data$fitted <- predict(model)
        
        original_data$resid <- residuals(model)
        
        
        
        #original_out[[j]][[i]] <- original_data
        #model_out[[j]][[i]] <- model
        
        
        ##residual vs fitted
        p1 <- ggplot(original_data, aes(fitted, resid)) + 
              geom_point() +
              geom_hline(yintercept = 0) +
              stat_smooth(method = "loess") +
              theme_minimal() +
              ggtitle(paste(j, i))
        
        p_vari[[j]][[i]] <- p1
        
        
        
        ##normality plot for the residuals
        p2 <- ggplot(original_data, aes(sample = resid)) +
              stat_qq() + stat_qq_line() +
              theme_minimal() +
              ggtitle(paste(j, i))
        
        p_qq[[j]][[i]] <- p2
        
      
        
        ##corr fitted vs observed
        p3 <- ggplot(original_data, aes(fitted, value)) +
              geom_point() +
              stat_smooth(method = "loess") + 
              theme_minimal() +
              ggtitle(paste(j, i))
        
        p_corr[[j]][[i]] <- p3
        
        
        library(forecast)
        p4 <- ggtsdisplay(original_data$resid)
        
        p_pacf[[j]][[i]] <- p4
      
      

  
  }
  
  
  

}



```

```{r}

library(gridExtra)

p_vari_s <- FlattenListofGgplot(p_vari)

p_qq_s <- FlattenListofGgplot(p_qq)

p_corr_s <- FlattenListofGgplot(p_corr)

p_pacf_s <- FlattenListofGgplot(p_pacf)



```

```{r}

rm(p1, p2, p3, p4, p_vari, p_vari_s, p_qq, p_qq_s, p_corr, p_corr_s, p_pacf, p_pacf_s, original_out, model_out,
   original_data, model)

```

```{r}

univar_mimo_apoE_e4_carrier_p_df <- data.frame(unlist(univar_mimo_p_apoE_e4_carrier))

#some imputations seem significant

univar_mimo_sampletime_p_df <- data.frame(unlist(univar_mimo_p_sample_time))

univar_mimo_interaction_p_df <- data.frame(unlist(univar_mimo_p_interaction))

univar_mimo_coef_apoE_e4_carrier_df <- data.frame(unlist(univar_mimo_coef_apoE_e4_carrier))

univar_mimo_coef_time_df <- data.frame(unlist(univar_mimo_coef_time))

univar_mimo_coef_interaction_df <- data.frame(unlist(univar_mimo_coef_interaction))

```


```{r}

univar_mimo_apoE_e4_carrier_p_df$meta <- rownames(univar_mimo_apoE_e4_carrier_p_df)

univar_mimo_sampletime_p_df$meta <- rownames(univar_mimo_sampletime_p_df)

univar_mimo_interaction_p_df$meta <- rownames(univar_mimo_interaction_p_df)

univar_mimo_coef_apoE_e4_carrier_df$meta <- rownames(univar_mimo_coef_apoE_e4_carrier_df)

univar_mimo_coef_time_df$meta <- rownames(univar_mimo_coef_time_df)

univar_mimo_coef_interaction_df$meta <- rownames(univar_mimo_coef_interaction_df)

```


```{r}

tables5_1 <- full_join(univar_mimo_coef_apoE_e4_carrier_df, univar_mimo_apoE_e4_carrier_p_df, by = "meta")

tables5_2 <- full_join(univar_mimo_coef_time_df, univar_mimo_sampletime_p_df, by = "meta")

tables5_3 <- full_join(univar_mimo_coef_interaction_df, univar_mimo_interaction_p_df, by = "meta")

tables5_4 <- full_join(tables5_1, tables5_2, by = "meta")

tables5 <- full_join(tables5_3, tables5_4, by = "meta")

rm(tables5_1, tables5_2, tables5_3, tables5_4)

```


```{r}

tables5 <- tables5 %>%
  separate(meta, into = c("olink_protein", "imputation_index"), sep = "\\.")

names(tables5)[1] <- "coef_interaction"
names(tables5)[2] #olink protein
names(tables5)[3] #imputation_index
names(tables5)[4] <- "p_interaction"
names(tables5)[5] <- "coef_apoeE4_carrier"
names(tables5)[6] <- "p_apoeE4_carrier"
names(tables5)[7] <- "coef_time_from_trauma"
names(tables5)[8] <- "p_time_from_trauma"

```

```{r}

tables5_summary <- tables5 %>%
  group_by(olink_protein) %>%
  summarise(across(c(coef_interaction, p_interaction:p_time_from_trauma), 
                   list(mean = ~ mean(.x, na.rm = T)))
            )


```

```{r}

tables5_summary$p_interaction_mean_adjusted <- p.adjust(tables5_summary$p_interaction_mean, method = "BH")

tables5_summary_sign <- tables5_summary %>%
  filter(p_interaction_mean <= 0.05)

View(tables5_summary_sign)

tables5_summary_sign <- tables5_summary_sign[, c(1:3, 8, 4:7)]



```



```{r}

rm(interim, interim2, 
   univar_mimo_apoE_e4_carrier_p_df, univar_mimo_apoE_e4_carrier_p_df_sign, univar_mimo_p_apoE_e4_carrier,
   i, j)

```


graphical illustration
```{r}

violin_data_interaction <- list()

olink_interaction_violin <- unique(tables5_summary_sign$olink_protein)

for(i in names(list_imputed_long)[1:20]){
  
  for(j in olink_interaction_violin[1:2]){
    
      interim <- list_imputed_long[[i]] %>%
        filter(olink_protein == j)
      
      violin_data_interaction[[j]][[i]] <- interim
      
      violin_data_interaction[[j]][[i]]$imputation_index <- i 
    
  }
  
}

```

```{r}

rm(interim)

```

```{r}

violin_merge_data_interaction <- list()
violin_interaction_plot <- list()

for(i in olink_interaction_violin[1:2]){
  
  interim <- bind_rows(violin_data_interaction[[i]])
  
  interim2 <- interim %>%
    group_by(Social.Security.Number, sample_time) %>%
    mutate(mean_value = mean(value, na.rm = T))
  
  interim2 <- interim2 %>%
    dplyr::select(-imputation_index)
  
  interim2 <- distinct(interim2)
  
  violin_merge_data_interaction[[i]] <- interim2

  interim_plot <- ggplot(interim2, aes(as.character(apoE_e4_carrier), mean_value, fill = sample_time)) +
  geom_violin(width = .75, alpha = .2, position = position_dodge(width = 0.75)) +
  geom_boxplot(width = .2, color = "grey", alpha = .4, position = position_dodge(width = 0.75)) +
  geom_point(pch = 21, size = 4, position = position_jitterdodge(jitter.width = .3)) +
  theme_linedraw() +
  xlab("APOE e4 carrier") + ylab("Protein quantification level") + ggtitle(paste0("OLINK protein: ", i))
  
  violin_interaction_plot[[i]] <- interim_plot
  
  
}


```


```{r}

library(cowplot)

grid <- plot_grid(violin_interaction_plot$CLM_1,
                  violin_interaction_plot$GFR_ALPHA_1,
                  nrow = 1)



```

prep data for additive model
```{r}

olink_protein_names_add <- olink_protein_names

```

remove redundant data
```{r}

rm(interim, interim_plot, interim2, univar_mimo, univar_mimo_apoE_e4_carrier_p_df, 
   univar_mimo_interaction_p_df, univar_mimo_interaction_p_df_sign, univar_mimo_p_apoE_e4_carrier,
   univar_mimo_p_interaction, univar_mimo_p_sample_time, violin_data_interaction, violin_interaction_plot,
   violin_merge_data_interaction, i, j, olink_interaction_violin, univar_mimo_sampletime_p_df)

```
MIXED MODELS (INTERACTION, adjusted for Age)
```{r}

library(nlme)

univar_mimo <- list()
univar_mimo_p_apoE_e4_carrier <- list()
univar_mimo_p_sample_time <- list()
univar_mimo_p_interaction <- list()
univar_mimo_p_age <- list()


for(i in names(list_imputed_long)[1:20]){
  
  for(j in olink_protein_names[1:90]){
    
    interim <- list_imputed_long[[i]] %>%
      filter(olink_protein == j)
    
    interim2 <- summary(lme(value ~ apoE_e4_carrier * sample_time + Age, 
                    random = ~1 | patientID_use,
                    data = interim))
    
    univar_mimo[[j]][[i]] <- interim2
    
    univar_mimo_p_apoE_e4_carrier[[j]][[i]] <- interim2$tTable[22]
    
    univar_mimo_p_sample_time[[j]][[i]] <- interim2$tTable[23]
    
    univar_mimo_p_age[[j]][[i]] <- interim2$tTable[24]
    
    univar_mimo_p_interaction[[j]][[i]] <- interim2$tTable[25]
    
  
  }
  
  
}


```

```{r}

model_out <- list()
original_out <- list()
p_vari <- list()
p_qq <- list()
p_corr <- list()
p_pacf <- list()


for(i in names(list_imputed_long)[1:20]){
  
  for(j in olink_protein_names[1:90]){

        original_data <- list_imputed_long[[i]] %>%
          filter(olink_protein == j)
        
        model <- lme(value ~ apoE_e4_carrier * sample_time + Age, 
                    random = ~1 | patientID_use,
                    data = original_data)
    

  
        original_data$fitted <- predict(model)
        
        original_data$resid <- residuals(model)
        

        
        ##residual vs fitted
        p1 <- ggplot(original_data, aes(fitted, resid)) + 
              geom_point() +
              geom_hline(yintercept = 0) +
              stat_smooth(method = "loess") +
              theme_minimal() +
              ggtitle(paste(j, i))
        
        p_vari[[j]][[i]] <- p1
        
        
        
        ##normality plot for the residuals
        p2 <- ggplot(original_data, aes(sample = resid)) +
              stat_qq() + stat_qq_line() +
              theme_minimal() +
              ggtitle(paste(j, i))
        
        p_qq[[j]][[i]] <- p2
        
      
        
        ##corr fitted vs observed
        p3 <- ggplot(original_data, aes(fitted, value)) +
              geom_point() +
              stat_smooth(method = "loess") + 
              theme_minimal() +
              ggtitle(paste(j, i))
        
        p_corr[[j]][[i]] <- p3
        
        library(forecast)
        p4 <- ggtsdisplay(original_data$resid)
        
        p_pacf[[j]][[i]] <- p4
      
      

  
  }
  
  
  

}



```

save all files
231021: this has been done previously and is therefore marked "#"
```{r}

library(gridExtra)

p_vari_s <- FlattenListofGgplot(p_vari)

p_qq_s <- FlattenListofGgplot(p_qq)

p_corr_s <- FlattenListofGgplot(p_corr)

p_pacf_s <- FlattenListofGgplot(p_pacf)


```

```{r}

rm(p1, p2, p3, p4, p_vari, p_vari_s, p_qq, p_qq_s, p_corr, p_corr_s, p_pacf, p_pacf_s, original_out, model_out,
   original_data, model)

```

```{r}

univar_mimo_apoE_e4_carrier_p_df <- data.frame(unlist(univar_mimo_p_apoE_e4_carrier))

#some imputations seem significant

univar_mimo_sampletime_p_df <- data.frame(unlist(univar_mimo_p_sample_time))

univar_mimo_interaction_p_df <- data.frame(unlist(univar_mimo_p_interaction))

univar_mimo_age_p_df <- data.frame(unlist(univar_mimo_p_age))

```

apoE e4
```{r}

univar_mimo_apoE_e4_carrier_p_df$meta <- rownames(univar_mimo_apoE_e4_carrier_p_df)

univar_mimo_apoE_e4_carrier_p_df <- univar_mimo_apoE_e4_carrier_p_df %>%
  separate(meta, into = c("olink_protein", "imputation_index"), sep = "\\.")

head(univar_mimo_apoE_e4_carrier_p_df)

names(univar_mimo_apoE_e4_carrier_p_df)[1] <- "p_apoe4_unadj"

univar_mimo_apoE_e4_carrier_p_df <- univar_mimo_apoE_e4_carrier_p_df %>%
  group_by(olink_protein) %>%
  summarise(p_apoe4_unadj_mean = mean(p_apoe4_unadj, na.rm = T))

univar_mimo_apoE_e4_carrier_p_df$p_apoe4_adj <- p.adjust(univar_mimo_apoE_e4_carrier_p_df$p_apoe4_unadj_mean, method = "BH")

univar_mimo_apoE_e4_carrier_p_df_sign <- univar_mimo_apoE_e4_carrier_p_df %>%
  filter(p_apoe4_unadj_mean <= 0.05)

univar_mimo_apoE_e4_carrier_p_df_sign



```

age
```{r}

univar_mimo_age_p_df$meta <- rownames(univar_mimo_age_p_df)

univar_mimo_age_p_df <- univar_mimo_age_p_df %>%
  separate(meta, into = c("olink_protein", "imputation_index"), sep = "\\.")

head(univar_mimo_age_p_df)

names(univar_mimo_age_p_df)[1] <- "p_age_unadj"

univar_mimo_age_p_df <- univar_mimo_age_p_df %>%
  group_by(olink_protein) %>%
  summarise(p_age_unadj_mean = mean(p_age_unadj, na.rm = T))

univar_mimo_age_p_df$p_age_adj <- p.adjust(univar_mimo_age_p_df$p_age_unadj_mean, method = "BH")

univar_mimo_age_p_df_sign <- univar_mimo_age_p_df %>%
  filter(p_age_unadj_mean <= 0.05)

univar_mimo_age_p_df_sign



```


```{r}

rm(interim, interim2, 
   univar_mimo_apoE_e4_carrier_p_df, univar_mimo_apoE_e4_carrier_p_df_sign, univar_mimo_p_apoE_e4_carrier,
   i, j,
   univar_mimo_age_p_df, univar_mimo_age_p_df_sign, univar_mimo_p_age)

```


interaction effect
```{r}

univar_mimo_interaction_p_df$meta <- rownames(univar_mimo_interaction_p_df)

univar_mimo_interaction_p_df <- univar_mimo_interaction_p_df %>%
  separate(meta, into = c("olink_protein", "imputation_index"), sep = "\\.")

head(univar_mimo_interaction_p_df)

names(univar_mimo_interaction_p_df)[1] <- "p_interaction_unadj"

univar_mimo_interaction_p_df <- univar_mimo_interaction_p_df %>%
  group_by(olink_protein) %>%
  summarise(p_interaction_unadj_mean = mean(p_interaction_unadj, na.rm = T))

univar_mimo_interaction_p_df$p_interaction_adj <- p.adjust(univar_mimo_interaction_p_df$p_interaction_unadj_mean, method = "BH") #do not use

univar_mimo_interaction_p_df_sign <- univar_mimo_interaction_p_df %>%
  filter(p_interaction_unadj_mean <= 0.05)

univar_mimo_interaction_p_df_sign



```

graphical illustration
```{r}

violin_data_interaction <- list()

olink_interaction_violin <- unique(univar_mimo_interaction_p_df_sign$olink_protein)

for(i in names(list_imputed_long)[1:20]){
  
  for(j in olink_interaction_violin[1:2]){
    
      interim <- list_imputed_long[[i]] %>%
        filter(olink_protein == j)
      
      violin_data_interaction[[j]][[i]] <- interim
      
      violin_data_interaction[[j]][[i]]$imputation_index <- i 
    
  }
  
}

```

```{r}

rm(interim)

```

```{r}

violin_merge_data_interaction <- list()
violin_interaction_plot <- list()

for(i in olink_interaction_violin[1:2]){
  
  interim <- bind_rows(violin_data_interaction[[i]])
  
  interim2 <- interim %>%
    group_by(Social.Security.Number, sample_time) %>%
    mutate(mean_value = mean(value, na.rm = T))
  
  interim2 <- interim2 %>%
    dplyr::select(-imputation_index)
  
  interim2 <- distinct(interim2)
  
  violin_merge_data_interaction[[i]] <- interim2

  interim_plot <- ggplot(interim2, aes(as.character(apoE_e4_carrier), mean_value, fill = sample_time)) +
  geom_violin(width = .75, alpha = .2, position = position_dodge(width = 0.75)) +
  geom_boxplot(width = .2, color = "grey", alpha = .4, position = position_dodge(width = 0.75)) +
  geom_point(pch = 21, size = 4, position = position_jitterdodge(jitter.width = .3)) +
  theme_linedraw() +
  xlab("APOE e4 carrier") + ylab("Protein quantification level") + ggtitle(paste0("OLINK protein: ", i))
  
  violin_interaction_plot[[i]] <- interim_plot
  
  
}


```


```{r}

library(cowplot)

grid <- plot_grid(violin_interaction_plot$CLM_1,
                  violin_interaction_plot$GFR_ALPHA_1,
                  nrow = 1)



```

prep data for additive model
```{r}

olink_protein_names_add <- olink_protein_names

```

remove redundant data
```{r}

rm(interim, interim_plot, interim2, univar_mimo, univar_mimo_apoE_e4_carrier_p_df, 
   univar_mimo_interaction_p_df, univar_mimo_interaction_p_df_sign, univar_mimo_p_apoE_e4_carrier,
   univar_mimo_p_interaction, univar_mimo_p_sample_time, violin_data_interaction, violin_interaction_plot,
   violin_merge_data_interaction, i, j, olink_interaction_violin, univar_mimo_sampletime_p_df)

```

MIXED MODELS (ADDITIVE)
```{r}

add_mimo <- list()
add_mimo_p_apoE_e4_carrier <- list()
add_mimo_p_sample_time <- list()


for(i in names(list_imputed_long)[1:20]){
  
  for(j in olink_protein_names_add[1:90]){
    
    interim <- list_imputed_long[[i]] %>%
      filter(olink_protein == j)
    
    interim2 <- summary(lme(value ~ apoE_e4_carrier + sample_time, 
                    random = ~1 | patientID_use,
                    data = interim))
    
    add_mimo[[j]][[i]] <- interim2
    
    add_mimo_p_apoE_e4_carrier[[j]][[i]] <- interim2$tTable[14]
    
    add_mimo_p_sample_time[[j]][[i]] <- interim2$tTable[15]
    
  
  }
  
  
}



```

```{r}

add_mimo_apoE_df <- data.frame(unlist(add_mimo_p_apoE_e4_carrier))
table(add_mimo_apoE_df$unlist.add_mimo_p_apoE_e4_carrier. <= 0.05) #T: 7

add_mimo_apoE_df$meta <- rownames(add_mimo_apoE_df)

add_mimo_apoE_df <- add_mimo_apoE_df %>%
  separate(meta, into = c("olink_protein", "imputation_index"), sep = "\\.")

head(add_mimo_apoE_df)

names(add_mimo_apoE_df)[1] <- "p_additive_unadj"

add_mimo_apoE_df <- add_mimo_apoE_df %>%
  group_by(olink_protein) %>%
  summarise(p_additive_unajd_mean = mean(p_additive_unadj, na.rm = T))

add_mimo_apoE_df$p_additive_adj <- p.adjust(add_mimo_apoE_df$p_additive_unajd_mean, method = "BH")

add_mimo_apoE_df_sign <- add_mimo_apoE_df %>%
  filter(p_additive_unajd_mean <= 0.05) #empty
  

####
add_mimo_sample_time_df <- data.frame(unlist(add_mimo_p_sample_time))
table(add_mimo_sample_time_df$unlist.add_mimo_p_sample_time. <= 0.05) #T914

add_mimo_sample_time_df$meta <- rownames(add_mimo_sample_time_df)

add_mimo_sample_time_df <- add_mimo_sample_time_df %>%
  separate(meta, into = c("olink_protein", "imputation_index"), sep = "\\.")
  
names(add_mimo_sample_time_df)[1] <- "p_additive_unadj"

add_mimo_sample_time_df <- add_mimo_sample_time_df %>%
  group_by(olink_protein) %>%
  summarise(p_additive_unadj_mean = mean(p_additive_unadj, na.rm = T))

add_mimo_sample_time_df$p_additive_adj <- p.adjust(add_mimo_sample_time_df$p_additive_unadj_mean, method = "BH")



```


assumptions
```{r}

model_out <- list()
original_out <- list()
p_vari <- list()
p_qq <- list()
p_corr <- list()
p_pacf <- list()


for(i in names(list_imputed_long)[1:20]){
  
  for(j in olink_protein_names[1:90]){

        original_data <- list_imputed_long[[i]] %>%
          filter(olink_protein == j)
        
        
        model <- lme(value ~ apoE_e4_carrier + sample_time, 
                    random = ~1 | patientID_use,
                    data = original_data)
    
  
        original_data$fitted <- predict(model)
        
        original_data$resid <- residuals(model)
        
        
        #original_out[[j]][[i]] <- original_data
        #model_out[[j]][[i]] <- model
        
        
        ##residual vs fitted
        p1 <- ggplot(original_data, aes(fitted, resid)) + 
              geom_point() +
              geom_hline(yintercept = 0) +
              stat_smooth(method = "loess") +
              theme_minimal() +
              ggtitle(paste(j, i))
        
        p_vari[[j]][[i]] <- p1
        
        
        
        ##normality plot for the residuals
        p2 <- ggplot(original_data, aes(sample = resid)) +
              stat_qq() + stat_qq_line() +
              theme_minimal() +
              ggtitle(paste(j, i))
        
        p_qq[[j]][[i]] <- p2
        
  
        
        ##corr fitted vs observed
        p3 <- ggplot(original_data, aes(fitted, value)) +
              geom_point() +
              stat_smooth(method = "loess") + 
              theme_minimal() +
              ggtitle(paste(j, i))
        
        p_corr[[j]][[i]] <- p3
        
       
        library(forecast)

        p4 <- ggtsdisplay(original_data$resid)
        
        p_pacf[[j]][[i]] <- p4
        
  
    
  }
  
  
  

}



```

save all files
```{r}

library(gridExtra)

p_vari_s <- FlattenListofGgplot(p_vari)

p_qq_s <- FlattenListofGgplot(p_qq)

p_corr_s <- FlattenListofGgplot(p_corr)

p_pacf_s <- FlattenListofGgplot(p_pacf)


```


```{r}

rm(p1, p2, p3, p4, p_vari, p_vari_s, p_qq, p_qq_s, p_corr, p_corr_s, p_pacf, p_pacf_s, original_out, model_out,
   original_data, model)

```

WE THUS PROCEED TO UNIVARIATE ASSESSMENTS

```{r}

rm(add_mimo, add_mimo_apoE_df, add_mimo_p_apoE_e4_carrier, add_mimo_sample_time_df,
   interim, interim2, univar_mimo_sampletime_p_df, i, j, inteirm_plot, violin_additive_plot,
   violin_data_additive, violin_data_interaction, violin_merge_data_additive, 
   olink_additive_violin)

```

MIXED MODELS (ADDITIVE)
```{r}

add_mimo <- list()
add_mimo_p_apoE_e4_carrier <- list()
add_mimo_p_sample_time <- list()
add_mimo_p_age <- list()


for(i in names(list_imputed_long)[1:20]){
  
  for(j in olink_protein_names_add[1:90]){
    
    interim <- list_imputed_long[[i]] %>%
      filter(olink_protein == j)
    
    interim2 <- summary(lme(value ~ apoE_e4_carrier + sample_time + Age, 
                    random = ~1 | patientID_use,
                    data = interim))
    
    add_mimo[[j]][[i]] <- interim2
    
    add_mimo_p_apoE_e4_carrier[[j]][[i]] <- interim2$tTable[18]
    
    add_mimo_p_sample_time[[j]][[i]] <- interim2$tTable[19]
    
    add_mimo_p_age[[j]][[i]] <- interim2$tTable[20]
    
  
  }
  
  
}



```

```{r}

add_mimo_apoE_df <- data.frame(unlist(add_mimo_p_apoE_e4_carrier))
table(add_mimo_apoE_df$unlist.add_mimo_p_apoE_e4_carrier. <= 0.05) #T: 7

add_mimo_apoE_df$meta <- rownames(add_mimo_apoE_df)

add_mimo_apoE_df <- add_mimo_apoE_df %>%
  separate(meta, into = c("olink_protein", "imputation_index"), sep = "\\.")

head(add_mimo_apoE_df)

names(add_mimo_apoE_df)[1] <- "p_additive_unadj"

add_mimo_apoE_df <- add_mimo_apoE_df %>%
  group_by(olink_protein) %>%
  summarise(p_additive_unajd_mean = mean(p_additive_unadj, na.rm = T))

add_mimo_apoE_df$p_additive_adj <- p.adjust(add_mimo_apoE_df$p_additive_unajd_mean, method = "BH")

add_mimo_apoE_df_sign <- add_mimo_apoE_df %>%
  filter(p_additive_unajd_mean <= 0.05) #empty
  

####
add_mimo_sample_time_df <- data.frame(unlist(add_mimo_p_sample_time))
table(add_mimo_sample_time_df$unlist.add_mimo_p_sample_time. <= 0.05) #T915

add_mimo_sample_time_df$meta <- rownames(add_mimo_sample_time_df)

add_mimo_sample_time_df <- add_mimo_sample_time_df %>%
  separate(meta, into = c("olink_protein", "imputation_index"), sep = "\\.")
  
names(add_mimo_sample_time_df)[1] <- "p_additive_unadj"

add_mimo_sample_time_df <- add_mimo_sample_time_df %>%
  group_by(olink_protein) %>%
  summarise(p_additive_unadj_mean = mean(p_additive_unadj, na.rm = T))

add_mimo_sample_time_df$p_additive_adj <- p.adjust(add_mimo_sample_time_df$p_additive_unadj_mean, method = "BH")


####
add_mimo_age_df <- data.frame(unlist(add_mimo_p_age))
table(add_mimo_age_df$unlist.add_mimo_p_age. <= 0.05) #T106

add_mimo_age_df$meta <- rownames(add_mimo_age_df)

add_mimo_age_df <- add_mimo_age_df %>%
  separate(meta, into = c("olink_protein", "imputation_index"), sep = "\\.")
  
names(add_mimo_age_df)[1] <- "p_additive_unadj"

add_mimo_age_df <- add_mimo_age_df %>%
  group_by(olink_protein) %>%
  summarise(p_additive_unadj_mean = mean(p_additive_unadj, na.rm = T))

add_mimo_age_df$p_additive_adj <- p.adjust(add_mimo_age_df$p_additive_unadj_mean, method = "BH")


```

assumptions
```{r}

model_out <- list()
original_out <- list()
p_vari <- list()
p_qq <- list()
p_corr <- list()
p_pacf <- list()


for(i in names(list_imputed_long)[1:20]){
  
  for(j in olink_protein_names[1:90]){

        original_data <- list_imputed_long[[i]] %>%
          filter(olink_protein == j)
        
        
        model <- lme(value ~ apoE_e4_carrier + sample_time + Age, 
                    random = ~1 | patientID_use,
                    data = original_data)
    
  
        original_data$fitted <- predict(model)
        
        original_data$resid <- residuals(model)
        

        
        ##residual vs fitted
        p1 <- ggplot(original_data, aes(fitted, resid)) + 
              geom_point() +
              geom_hline(yintercept = 0) +
              stat_smooth(method = "loess") +
              theme_minimal() +
              ggtitle(paste(j, i))
        
        p_vari[[j]][[i]] <- p1
        
        
        
        ##normality plot for the residuals
        p2 <- ggplot(original_data, aes(sample = resid)) +
              stat_qq() + stat_qq_line() +
              theme_minimal() +
              ggtitle(paste(j, i))
        
        p_qq[[j]][[i]] <- p2
        
  
        
        ##corr fitted vs observed
        p3 <- ggplot(original_data, aes(fitted, value)) +
              geom_point() +
              stat_smooth(method = "loess") + 
              theme_minimal() +
              ggtitle(paste(j, i))
        
        p_corr[[j]][[i]] <- p3
        
       
        library(forecast)

        p4 <- ggtsdisplay(original_data$resid)
        
        p_pacf[[j]][[i]] <- p4
        
  
    
  }
  
  
  

}



```

save all files
```{r}

library(gridExtra)

p_vari_s <- FlattenListofGgplot(p_vari)

p_qq_s <- FlattenListofGgplot(p_qq)

p_corr_s <- FlattenListofGgplot(p_corr)

p_pacf_s <- FlattenListofGgplot(p_pacf)



```


```{r}

rm(p1, p2, p3, p4, p_vari, p_vari_s, p_qq, p_qq_s, p_corr, p_corr_s, p_pacf, p_pacf_s, original_out, model_out,
   original_data, model)

```

WE THUS PROCEED TO UNIVARIATE ASSESSMENTS

```{r}

rm(add_mimo, add_mimo_apoE_df, add_mimo_p_apoE_e4_carrier, add_mimo_sample_time_df,
   interim, interim2, univar_mimo_sampletime_p_df, i, j, inteirm_plot, violin_additive_plot,
   violin_data_additive, violin_data_interaction, violin_merge_data_additive, 
   olink_additive_violin,
   add_mimo_age_df, add_mimo_p_age)

```



MIXED MODELS (UNIVARIATE: APOE E4)
```{r}

apoe_mimo <- list()
apoe_mimo_p <- list()


for(i in names(list_imputed_long)[1:20]){
  
  for(j in olink_protein_names[1:90]){
    
    interim <- list_imputed_long[[i]] %>%
      filter(olink_protein == j)
    
    interim2 <- summary(lme(value ~ apoE_e4_carrier, 
                    random = ~1 | patientID_use,
                    data = interim))
    
    apoe_mimo[[j]][[i]] <- interim2
    
    apoe_mimo_p[[j]][[i]] <- interim2$tTable[10]
  
  }
  
  
}



```

```{r}

apoe_mimo_df <- data.frame(unlist(apoe_mimo_p))

table(apoe_mimo_df$unlist.apoe_mimo_p. <= 0.05) #T: 7 

```

new 231013 starts
new 231012 - generalized solution reg assumptions
```{r}

model_out <- list()
original_out <- list()
p_vari <- list()
p_qq <- list()
p_corr <- list()
p_pacf <- list()


for(i in names(list_imputed_long)[1:20]){
  
  for(j in olink_protein_names[1:90]){

        original_data <- list_imputed_long[[i]] %>%
          filter(olink_protein == j)
        
        
        model <- lme(value ~ apoE_e4_carrier, 
                    random = ~1 | patientID_use,
                    data = original_data)
    
  
        original_data$fitted <- predict(model)
        
        original_data$resid <- residuals(model)
        
        
        
        #original_out[[j]][[i]] <- original_data
        #model_out[[j]][[i]] <- model
        
        
        ##residual vs fitted
        p1 <- ggplot(original_data, aes(fitted, resid)) + 
              geom_point() +
              geom_hline(yintercept = 0) +
              stat_smooth(method = "loess") +
              theme_minimal() +
              ggtitle(paste(j, i))
        
        p_vari[[j]][[i]] <- p1
        
        
        
        ##normality plot for the residuals
        p2 <- ggplot(original_data, aes(sample = resid)) +
              stat_qq() + stat_qq_line() +
              theme_minimal() +
              ggtitle(paste(j, i))
        
        p_qq[[j]][[i]] <- p2
        
  
        
        ##corr fitted vs observed
        p3 <- ggplot(original_data, aes(fitted, value)) +
              geom_point() +
              stat_smooth(method = "loess") + 
              theme_minimal() +
              ggtitle(paste(j, i))
        
        p_corr[[j]][[i]] <- p3
        
       
        library(forecast)

        p4 <- ggtsdisplay(original_data$resid)
        
        p_pacf[[j]][[i]] <- p4
        
  
    
  }
  
  
  

}



```


save all files
```{r}

library(gridExtra)

p_vari_s <- FlattenListofGgplot(p_vari)

p_qq_s <- FlattenListofGgplot(p_qq)

p_corr_s <- FlattenListofGgplot(p_corr)

p_pacf_s <- FlattenListofGgplot(p_pacf)



```


```{r}

rm(p1, p2, p3, p4, p_vari, p_vari_s, p_qq, p_qq_s, p_corr, p_corr_s, p_pacf, p_pacf_s, original_out, model_out,
   original_data, model)

```


```{r}

apoe_mimo_df$meta <- rownames(apoe_mimo_df)

apoe_mimo_df <- apoe_mimo_df %>%
  separate(meta, into = c("olink_protein", "imputation_index"), sep = "\\.")

head(apoe_mimo_df)

names(apoe_mimo_df)[1] <- "p_univar_unadj"

apoe_mimo_df <- apoe_mimo_df %>%
  group_by(olink_protein) %>%
  summarise(p_univar_unadj_mean = mean(p_univar_unadj, na.rm = T))

apoe_mimo_df$p_univar_adj <- p.adjust(apoe_mimo_df$p_univar_unadj_mean, method = "BH")

apoe_mimo_df_sign <- apoe_mimo_df %>%
  filter(p_univar_unadj_mean <= 0.05) 

apoe_mimo_df_sign #empty


```



MIXED MODELS (UNIVARIATE: TIME)
```{r}

time_mimo <- list()
time_mimo_p <- list()
time_mimo_t <- list()

for(i in names(list_imputed_long)[1:20]){
  
  for(j in olink_protein_names[1:90]){
    
    interim <- list_imputed_long[[i]] %>%
      filter(olink_protein == j)
    
    interim2 <- summary(lme(value ~ sample_time, 
                    random = ~1 | patientID_use,
                    data = interim))
    
    time_mimo[[j]][[i]] <- interim2
    
    time_mimo_p[[j]][[i]] <- interim2$tTable[10]
    
    time_mimo_t[[j]][[i]] <- interim2$tTable[8]
  
  }
  
  
}



```


```{r}

model_out <- list()
original_out <- list()
p_vari <- list()
p_qq <- list()
p_corr <- list()
p_pacf <- list()


for(i in names(list_imputed_long)[1:20]){
  
  for(j in olink_protein_names[1:90]){

        original_data <- list_imputed_long[[i]] %>%
          filter(olink_protein == j)
        
        
        model <- lme(value ~ sample_time, 
                    random = ~1 | patientID_use,
                    data = original_data)

  
        original_data$fitted <- predict(model)
        
        original_data$resid <- residuals(model)
        
        

        ##residual vs fitted
        p1 <- ggplot(original_data, aes(fitted, resid)) + 
              geom_point() +
              geom_hline(yintercept = 0) +
              stat_smooth(method = "loess") +
              theme_minimal() +
              ggtitle(paste(j, i))
        
        p_vari[[j]][[i]] <- p1
        
        
        
        ##normality plot for the residuals
        p2 <- ggplot(original_data, aes(sample = resid)) +
              stat_qq() + stat_qq_line() +
              theme_minimal() +
              ggtitle(paste(j, i))
        
        p_qq[[j]][[i]] <- p2
        
          
        
        ##corr fitted vs observed
        p3 <- ggplot(original_data, aes(fitted, value)) +
              geom_point() +
              stat_smooth(method = "loess") + 
              theme_minimal() +
              ggtitle(paste(j, i))
        
        p_corr[[j]][[i]] <- p3
        
        
        library(forecast)

        p4 <- ggtsdisplay(original_data$resid)
        
        p_pacf[[j]][[i]] <- p4
        
  
    
  }
  
  
  

}



```


save all files
```{r}

library(gridExtra)

p_vari_s <- FlattenListofGgplot(p_vari)

p_qq_s <- FlattenListofGgplot(p_qq)

p_corr_s <- FlattenListofGgplot(p_corr)

p_pacf_s <- FlattenListofGgplot(p_pacf)



```

```{r}

rm(p1, p2, p3, p4, p_vari, p_vari_s, p_qq, p_qq_s, p_corr, p_corr_s, p_pacf, p_pacf_s, original_out, model_out,
   original_data, model)

```


```{r}

time_mimo_df <- data.frame(unlist(time_mimo_p))
table(time_mimo_df$unlist.time_mimo_p. <= 0.05) 
time_mimo_df$meta <- rownames(time_mimo_df)


time_mimo_df_t <- data.frame(unlist(time_mimo_t))
time_mimo_df_t$meta <- rownames(time_mimo_df_t)

time_mimo_df <- full_join(time_mimo_df, time_mimo_df_t, by = "meta")


```



```{r}


time_mimo_df <- time_mimo_df %>%
  separate(meta, into = c("olink_protein", "imputation_index"), sep = "\\.")

head(time_mimo_df)

names(time_mimo_df)[1] <- "p_time_unadj"
names(time_mimo_df)[4] <- "t_value"

time_mimo_df <- time_mimo_df %>%
  group_by(olink_protein) %>%
  summarise(p_time_unadj_mean = mean(p_time_unadj, na.rm = T),
            t_time_mean = mean(t_value, na.rm = T))

time_mimo_df$p_time_adj <- p.adjust(time_mimo_df$p_time_unadj_mean, method = "BH") 

time_mimo_df_sign <- time_mimo_df %>%
  filter(p_time_unadj_mean <= 0.05)

time_mimo_df_sign <- time_mimo_df_sign[, c(1, 3, 2, 4)]


length(unique(time_mimo_df_sign$olink_protein[time_mimo_df_sign$p_time_unadj_mean <= 0.05])) #41
length(unique(time_mimo_df_sign$olink_protein[time_mimo_df_sign$p_time_adj <= 0.05])) #34


```


compare against OLD time dependent vars
```{r}

old_univar_time_mimo_df <- read.csv("./output/draft/regressions/univar_time_231031_CL.csv")

old_univar_time_mimo_df <- old_univar_time_mimo_df[, 2:4]

old_univar_time_mimo_df_sign <- old_univar_time_mimo_df %>%
  filter(p_time_unadj_mean <= 0.05)

setdiff(time_mimo_df_sign$olink_protein, old_univar_time_mimo_df_sign$olink_protein) #[1] "Abeta_40" "DKK_4"  (unique in the new df) 
setdiff(old_univar_time_mimo_df_sign$olink_protein, time_mimo_df_sign$olink_protein) #empty

#hence - two more proteins are added following the new imputations


```

graphical illustration 
```{r}

violin_data_time <- list()

olink_time_violin <- unique(time_mimo_df_sign$olink_protein)

for(i in names(list_imputed_long)[1:20]){
  
  for(j in olink_time_violin[1:41]){
      #note length(olink_time_violin) == 41 with updated imputations
    
      interim <- list_imputed_long[[i]] %>%
        filter(olink_protein == j)
      
      violin_data_time[[j]][[i]] <- interim
      
      violin_data_time[[j]][[i]]$imputation_index <- i 
    
  }
  
}



```

```{r}
rm(interim, interim2)
```

```{r}

violin_merge_data_time <- list()
violin_time_plot <- list()

for(i in olink_time_violin[1:41]){
  
  interim <- bind_rows(violin_data_time[[i]])
  
  interim2 <- interim %>%
    group_by(Social.Security.Number, sample_time) %>%
    mutate(mean_value = mean(value, na.rm = T))
  
  interim2 <- interim2 %>%
    dplyr::select(-c(imputation_index, value))
  
  interim2 <- distinct(interim2)
  
  violin_merge_data_time[[i]] <- interim2

  interim_plot <- ggplot(interim2, aes(sample_time, mean_value, fill = as.factor(sample_time))) +
  geom_violin(width = .75, alpha = .2, position = position_dodge(width = 0.75)) +
  geom_boxplot(width = .2, color = "grey", alpha = .4, position = position_dodge(width = 0.75)) +
  geom_point(pch = 21, size = 4, position = position_jitterdodge(jitter.width = .3)) +
  scale_fill_viridis_d() +
  theme_linedraw() +
  xlab("Sample time") + ylab("Protein quantification level") + ggtitle(paste0("OLINK protein: ", i))
  
  violin_time_plot[[i]] <- interim_plot
  
  
}

```

```{r}

rm(interim, interim2)

```

```{r}

violin_merge_data_time_supp <- list()
violin_time_plot_supp <- list()

for(i in olink_time_violin[1:41]){
  
  interim_supp <- bind_rows(violin_data_time[[i]])
  
  interim2_supp <- interim_supp %>%
    group_by(Social.Security.Number, sample_time) %>%
    mutate(mean_value = mean(value, na.rm = T))
  
  interim2_supp <- interim2_supp %>%
    dplyr::select(-c(imputation_index, value))
  
  interim2_supp <- distinct(interim2_supp)
  
  interim3_supp <- interim2_supp %>%
  filter(sample_time == "d1") %>%
  mutate(day1_value = mean_value) %>%
  dplyr::select(c(patientID_use, day1_value))
  
  interim2_supp <- left_join(interim2_supp, interim3_supp, by = "patientID_use")
  
  interim2_supp <- interim2_supp %>%
  dplyr::select(-c(Social.Security.Number.y, sample_time.y))
  
  names(interim2_supp)[names(interim2_supp) == "sample_time.x"] <- "sample_time"
  
  interim2_supp$value_norm <- interim2_supp$mean_value/interim2_supp$day1_value

  interim2_supp$sample_time_numeric <- if_else(interim2_supp$sample_time == "d1", 1, 3)
  
  violin_merge_data_time_supp[[i]] <- interim2_supp

  interim_plot_supp <- ggplot(interim2_supp, aes(sample_time_numeric, value_norm)) +
  geom_point(pch = 21, size = 4, fill = "gray") +
  geom_line(aes(group = as.factor(patientID_use)), col = "gray") + 
  geom_smooth(method = "loess", col = "#440154", se = F) +
  labs(x = "Sample time",
       y = "Normalized protein concentration") +
  ggtitle(paste0("OLINK protein: ", i)) +
  theme_minimal()
  
  violin_time_plot_supp[[i]] <- interim_plot_supp
  
  
}


time_plot_supp <- FlattenListofGgplot(violin_time_plot_supp)




```


```{r}
rm(interim, interim_plot, interim2, 
   time_mimo, time_mimo_df, time_mimo_df_sign, time_mimo_p, 
   violin_data_time, violin_merge_data_time, violin_time_plot,
   i, j, olink_protein_names_add, olink_time_violin,
   grid, add_mimo_apoE_df, add_mimo_apoE_df_sign, add_mimo_p_sample_time,
   apoe_mimo, apoe_mimo_df, apoe_mimo_df_sign, apoe_mimo_p)

```


graphical illustration neurodegenerative proteins
231208: updated to also include MAPT

```{r}

violin_data_time <- list()

olink_neurodegen_violin <- c("Abeta_38", "Abeta_40", "Abeta_42", "Abeta42_40_ratio", "MAPT")

for(i in names(list_imputed_long)[1:20]){
  
  for(j in olink_neurodegen_violin[1:5]){
    
      interim <- list_imputed_long[[i]] %>%
        filter(olink_protein == j)
      
      violin_data_time[[j]][[i]] <- interim
      
      violin_data_time[[j]][[i]]$imputation_index <- i 
    
  }
  
}



```


```{r}
rm(interim, i, j)
```

```{r}

violin_merge_data_time <- list()
violin_time_plot <- list()

for(i in olink_neurodegen_violin[1:5]){
  
  interim <- bind_rows(violin_data_time[[i]])
  
  interim2 <- interim %>%
    group_by(Social.Security.Number, sample_time) %>%
    mutate(mean_value = mean(value, na.rm = T))
  
 interim2 <- interim2 %>%
   dplyr::select(-c(imputation_index, value))
 
 interim2 <- distinct(interim2)
 
 violin_merge_data_time[[i]] <- interim2

 interim_plot <- ggplot(interim2, aes(as.character(apoE_e4_carrier), mean_value, fill = as.factor(sample_time))) +
 geom_violin(width = .75, alpha = .2, position = position_dodge(width = 0.75)) +
 geom_boxplot(width = .2, color = "grey", alpha = .4, position = position_dodge(width = 0.75)) +
 geom_point(pch = 21, size = 4, position = position_jitterdodge(jitter.width = .3)) +
 scale_fill_viridis_d() +
 theme_linedraw() +
 xlab("APOE e4 carrier") + ylab("Protein quantification level") + ggtitle(paste0("OLINK protein: ", i))
 
 violin_time_plot[[i]] <- interim_plot
  
  
}

```


```{r}

print(Sys.time())
print("all code run")


```




